---
title: "W4 - Data Cleaning"
author: "atorbati"
date: "2026-02-02"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```
```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE,
  message = FALSE,
  warning = FALSE
)
```

## 1) Project topic and dataset description

This project examines longitudinal trends in U.S. suicide mortality rates and how rates differ by demographic group (sex, race/ethnicity, age) using national-level rate estimates.

Each row represents one national estimate of the suicide death rate per 100,000 resident population for a specific year, age group, and demographic subgroup defined by the stratification fields (`STUB_NAME` + `STUB_LABEL`). The `FLAG` field indicates N/A (ellipses) or unreliable estimates (asterisk, often due to small counts).

## 2) Read data

```{r read-data}
library(readr)
library(dplyr)
library(tidyr)
library(ggplot2)
library(knitr)
library(stringr)

file_path <- "/Users/user/Downloads/Death_rates_for_suicide__by_sex__race__Hispanic_origin__and_age__United_States (1).csv"
# If you prefer not to hardcode paths, comment above and use:
# file_path <- file.choose()

stopifnot(file.exists(file_path))

suicide_raw <- read_csv(file_path, show_col_types = FALSE)

# small, safe previews (do NOT print whole data)
dim(suicide_raw)
head(suicide_raw, 8)
```

## 3) Prioritize variables for research (minimum 10 variables)

Your questions focus on:
- Time: YEAR
- Age groups: AGE (categorical) + AGE_NUM (ordering code)
- Demographic stratification: STUB_NAME / STUB_LABEL (and their numeric codes)
- Outcome: ESTIMATE (rate)
- Data quality: FLAG

To meet the requirement (≥10 variables; ≥6 numeric; ≥2 categorical), I will keep exactly 10 variables:
- Numeric (6): UNIT_NUM, STUB_NAME_NUM, STUB_LABEL_NUM, YEAR, AGE_NUM, ESTIMATE
- Categorical (4): STUB_NAME, STUB_LABEL, AGE, FLAG

Before cleaning, I also check unit consistency and basic structure.

```{r structure-checks}
# column types
type_table <- tibble(
  column = names(suicide_raw),
  class  = sapply(suicide_raw, function(x) paste(class(x), collapse = ", "))
)
kable(type_table, caption = "Column types (raw data)")

# key data quality checks
data.frame(
  total_rows = nrow(suicide_raw),
  duplicate_rows = sum(duplicated(suicide_raw)),
  missing_estimate = sum(is.na(suicide_raw$ESTIMATE)),
  missing_flag = sum(is.na(suicide_raw$FLAG))
) %>% kable(caption = "Quality checks (raw data)")
```

## 4) Clean your data

### 4.1 Units: prevent mixing incompatible measurements

Some datasets include multiple units (e.g., crude vs age-adjusted). I will choose one unit and filter to it.

```{r unit-values}
unit_values <- sort(unique(suicide_raw$UNIT))
unit_values
```

```{r choose-unit}
# Prefer an age-adjusted "Deaths per 100,000" unit if present; otherwise keep the first unit.
unit_to_keep <- unit_values[
  which(str_detect(tolower(unit_values), "death") &
        str_detect(tolower(unit_values), "100,000"))
][1]

if (is.na(unit_to_keep)) {
  unit_to_keep <- unit_values[1]
}

unit_to_keep
```

### 4.2 YEAR vs YEAR_NUM redundancy

If YEAR and YEAR_NUM match exactly, YEAR_NUM is redundant.

```{r year-redundancy-check}
year_matches <- all(suicide_raw$YEAR == suicide_raw$YEAR_NUM, na.rm = TRUE)
year_matches
```

### 4.3 Apply cleaning rules

Cleaning actions:
1. Filter to one unit (UNIT == unit_to_keep)
2. Remove exact duplicates (distinct())
3. Remove rows where ESTIMATE is missing (cannot analyze missing outcomes)
4. Remove N/A rows indicated by FLAG == "..." (not applicable)
5. Do NOT auto-remove “outliers” for YEAR or ESTIMATE (extremes can be real; YEAR is not an outlier concept)
6. Drop YEAR_NUM if redundant
7. Keep the prioritized 10 variables (6 numeric + 4 categorical)
(NOTE: understand that some of the variables are a bit repetitve - if this is an issue please communicate this with me via EMAIL and let me know if a different dataset is needed.)

```{r clean-data}
suicide_clean <- suicide_raw %>%
  filter(UNIT == unit_to_keep) %>%
  distinct() %>%
  filter(!is.na(ESTIMATE)) %>%
  filter(is.na(FLAG) | FLAG != "...") %>%
  { if (year_matches) select(., -YEAR_NUM) else . } %>%
  select(
    # categorical (4)
    STUB_NAME, STUB_LABEL, AGE, FLAG,
    # numeric (6)
    UNIT_NUM, STUB_NAME_NUM, STUB_LABEL_NUM, YEAR, AGE_NUM, ESTIMATE
  )

dim(suicide_clean)
head(suicide_clean, 8)
```

### 4.4 Confirm we meet the “minimum variables” rule and types

```{r confirm-variables}
data.frame(
  n_variables = ncol(suicide_clean),
  n_numeric = sum(sapply(suicide_clean, is.numeric)),
  n_categorical = sum(sapply(suicide_clean, function(x) is.character(x) || is.factor(x)))
) %>% kable(caption = "Variable count check (cleaned data)")

kable(
  tibble(variable = names(suicide_clean),
         class = sapply(suicide_clean, function(x) paste(class(x), collapse = ", "))),
  caption = "Column types (cleaned data)"
)
```

## 5) Summary statistics (cleaned data)

### 5.1 Numeric summary table (≥4 numeric variables)

ESTIMATE is the main outcome (rate). YEAR is time. AGE_NUM and STUB_LABEL_NUM are numeric ordering/coding variables that support sorting and subgroup structure.

```{r numeric-summary}
num_vars <- c("ESTIMATE", "YEAR", "AGE_NUM", "STUB_LABEL_NUM")

numeric_summary <- suicide_clean %>%
  summarise(across(all_of(num_vars),
                   list(
                     min  = ~ min(., na.rm = TRUE),
                     max  = ~ max(., na.rm = TRUE),
                     mean = ~ mean(., na.rm = TRUE),
                     sd   = ~ sd(., na.rm = TRUE)
                   ),
                   .names = "{.col}__{.fn}")) %>%
  pivot_longer(everything(), names_to = "metric", values_to = "value") %>%
  separate(metric, into = c("variable","stat"), sep = "__") %>%
  pivot_wider(names_from = stat, values_from = value) %>%
  select(variable, min, max, mean, sd)

kable(numeric_summary, digits = 3, caption = "Numeric summary statistics (cleaned data)")
```

### 5.2 Simple plot (histogram)

```{r histogram-estimate}
ggplot(suicide_clean, aes(x = ESTIMATE)) +
  geom_histogram(bins = 30) +
  labs(
    title = "Histogram of ESTIMATE (cleaned data)",
    x = "ESTIMATE (rate)",
    y = "Count"
  )
```

### 5.3 Categorical summaries (≥2 categorical variables)

For each categorical variable, I report:
- number of unique values
- frequency table (top levels only as the rest aren't as significant)

#### Categorical variable 1: AGE

```{r cat-age}
n_unique_age <- n_distinct(suicide_clean$AGE)
n_unique_age

age_freq <- suicide_clean %>%
  count(AGE, sort = TRUE)

kable(head(age_freq, 15), caption = "AGE frequency table (top 15)")
```

#### Categorical variable 2: STUB_LABEL

```{r cat-stub-label}
n_unique_stub <- n_distinct(suicide_clean$STUB_LABEL)
n_unique_stub

stub_freq <- suicide_clean %>%
  count(STUB_LABEL, sort = TRUE)

kable(head(stub_freq, 15), caption = "STUB_LABEL frequency table (top 15)")
```

(Additional categorical checks, not required but useful)

```{r cat-stub-name}
n_unique_stubname <- n_distinct(suicide_clean$STUB_NAME)
n_unique_stubname

stubname_freq <- suicide_clean %>%
  count(STUB_NAME, sort = TRUE)

kable(head(stubname_freq, 15), caption = "STUB_NAME frequency table (top 15)")
```

## 6) Research questions 

These questions come from my initial proposal (and are still appropriate after cleaning). I refined wording slightly to match the cleaned structure (YEAR, AGE, STUB fields, ESTIMATE).

1. Disparity: How do suicide death rates differ by sex and race/ethnicity, and have these disparities widened or narrowed over time?

2. Age: Which age groups are the most affected by suicide, and have the highest-risk age groups changed over time?

3. Sex: Are men or women more impacted by suicide, and does the sex difference persist across age groups?

## 7) Planned plots for next assignment...

### For Q1 (sex/race disparity over time)
Planned plots:
- Line chart: YEAR (x) vs ESTIMATE (y), color = subgroup label (STUB_LABEL), filtered to a specific stratification scheme (STUB_NAME) such as sex or race/ethnicity.
- Optional: Gap-over-time plot (difference between two groups per year). (longitudinal maybe?)

### For Q2 (age groups most affected and change over time)
Planned plots:
- Heatmap: YEAR (x) by AGE (y) with fill = ESTIMATE to detect shifting risk across decades.
- Small-multiples line plots: ESTIMATE over time for selected age groups.

### For Q3 (sex differences across age groups)
Planned plots:
- Facet line chart: YEAR vs ESTIMATE, color by sex subgroup, faceted by AGE (or a subset of age bands).
- Optional: Boxplots of ESTIMATE by sex within chosen age ranges/time windows. (boxplots are great to see the IQR and visualize outliers if needed to construct an argument.)

PS:
Thank you for reading - if you have ANY questions or concerns (if my writing is not professional enough, need to reword certain sections, etc.) please contact me via email. I really care about this class and I want to get a good grade to show to my mentor that taught me R at RAND. :) I worked on a suicide research project across Pittsburgh collecting data so this is very important to me. 